---
title: "1st year Atelier Statistique - Mentored by Mr. Samer Azzabi"
output: html_notebook
---

# Title of the Subject
Analysis of Credit Card Transactions for Early Detection of Defaults.

# Subject Description
This study focuses on the in-depth analysis of credit card holders' behavior with the aim of early detection of potential signs of payment default. By examining spending habits, repayment patterns, and other financial behaviors, the objective is to identify predictive indicators that enable the assessment of default risk. The goal of this research is to facilitate a proactive approach in managing payment defaults, thereby offering financial institutions more accurate and faster solutions to evaluate and mitigate the inherent risks associated with credit card transactions.
Work to be Done

• Data preprocessing (Outliers, missing values, data quality analysis)
• Exploratory data analysis (Univariate analysis, bivariate analysis, Data visualization, ...)
• Variable transformation (Normalization, discretization, ...)
• Correlation analysis, analysis of variance, independence tests
• Dashboard with RShiny.

# Information About the Dataset:
The dataset used is the "Default of Credit Card Clients Dataset" from UCI ML on Kaggle. This dataset contains information on payment defaults, demographic factors, credit data, payment history, and billing statements of credit card clients in Taiwan from April 2005 to September 2005.


```{r}
#Importing dependencies
library(ggplot2)
library(moments)
library(patchwork)
```

```{r}
# Importing Data
data <- read.csv("./UCI_Credit_Card.csv")
head(data)
```
# Content
• ID: ID of each client
• LIMIT_BAL: Amount of credit granted in NT dollars (includes individual and family/supplementary credit)
• SEX: Gender (1=male, 2=female)
• EDUCATION: (1=graduate school, 2=university, 3=high school, 4=others, 5=unknown, 6=unknown)
• AGE: Age in years
• PAY_0: Repayment status in September 2005 (-1=pay duly, 1=payment delay of one month, 2=payment delay of two months, ... 8=payment delay of eight months, 9=payment delay of nine months and more)
• PAY_2: Repayment status in August 2005 (scale identical to the one above)
• PAY_3: Repayment status in July 2005 (scale identical to the one above)
• PAY_4: Repayment status in June 2005 (scale identical to the one above)
• PAY_5: Repayment status in May 2005 (scale identical to the one above)
• PAY_6: Repayment status in April 2005 (scale identical to the one above)
• BILL_AMT1: Bill statement amount in September 2005 (NT dollar)
• BILL_AMT2: Bill statement amount in August 2005 (NT dollar)
• BILL_AMT3: Bill statement amount in July 2005 (NT dollar)
• BILL_AMT4: Bill statement amount in June 2005 (NT dollar)
• BILL_AMT5: Bill statement amount in May 2005 (NT dollar)
• BILL_AMT6: Bill statement amount in April 2005 (NT dollar)
• PAY_AMT1: Previous payment amount in September 2005 (NT dollar)
• PAY_AMT2: Previous payment amount in August 2005 (NT dollar)
• PAY_AMT3: Previous payment amount in July 2005 (NT dollar)
• PAY_AMT4: Previous payment amount in June 2005 (NT dollar)
• PAY_AMT5: Previous payment amount in May 2005 (NT dollar)
• PAY_AMT6: Previous payment amount in April 2005 (NT dollar)
• default.payment.next.month: Default payment (1=yes, 0=no)

# Structural check
```{r}
# Shape
print(paste("Shape:", paste(dim(data), collapse = " x ")))

# Data types
print("Data Types:")
sapply(data, class)

# Memory
print(paste("Data usage:", format(object.size(data), units = "MB"), collapse = "x"))
```
```{r}
# Total of missin values
sum(is.na(data))
```
It appears that the data doesn't contain any missing values, we'll return to that later, let's check for duplicate values
```{r}
# Checking for duplicate data
any(duplicated(data))
```
It seems the data doesn't contain any duplicate rows


# Defining variable types

```{r}
data <- subset(data, select = -ID)
factor_vars <- c("SEX", "EDUCATION", "MARRIAGE", "PAY_0", "PAY_2", "PAY_3", "PAY_4",
                "PAY_5", "PAY_6")
quant_vars <- c("LIMIT_BAL", "AGE", "BILL_AMT1", "BILL_AMT2", "BILL_AMT3",
                "BILL_AMT4", "BILL_AMT5", "BILL_AMT6")
data[factor_vars] <- lapply(data[factor_vars], function(x) as.factor(x))
data$default.payment.next.month <- factor(
  data$default.payment.next.month,
  levels = c(0, 1),
  labels = c("Non-Defaut", "Defaut")
)
target <-"default.payment.next.month"
```

# Univariate Analysis
```{r}
# Function to detect outliers using IQR
detect_outliers_iqr <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(x < lower_bound | x > upper_bound)
}
```
Let's take a look at our variables one by one.
```{r}
ggplot(data, aes(x = LIMIT_BAL)) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 30, fill = "steelblue", alpha = 0.7) +
    geom_density(color = "red", linewidth = 1) +
    labs(title = "LIMIT_BAL", x = NULL, y = NULL) +
    theme_minimal()
```
We notice a positive skewness: most clients have credit limits below 500000NTD, while very few have higher credit limits.
```{r}
LIMIT_BAL_skewness = skewness(data$LIMIT_BAL)
round(LIMIT_BAL_skewness, digits = 3)
```
The skewness value is 0.993 which exists in the interval [0.5, 1], this confirms that the data is moderately positively skewed.
```{r}
ggplot(data, aes(x = LIMIT_BAL)) +
    geom_boxplot() +
    theme_classic()
```
Notice the individuals we mentioned earlier as the few who have a credit limit of over 500000NTD are what consists our outliers in the case of LIMIT_BAL. Let's take a closer look at these outliers:
```{r}
outliers = detect_outliers_iqr(data$LIMIT_BAL)
print(paste("Outlier count:", sum(outliers)))
print(paste("Outlier percentage:", sum(outliers)/nrow(data) * 100))
```
So only 0.5% of the population has a credit limit of over 500000NTD, which is a healthy percentage of outliers.

```{r, fig.width=14, fig.height=15}
# Create a list of plots
plots <- lapply(quant_vars, function(var) {
  ggplot(data, aes(x = .data[[var]])) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 30, fill = "steelblue", alpha = 0.7) +
    geom_density(color = "red", linewidth = 1) +
    labs(title = var, x = NULL, y = NULL) +
    theme_minimal()
})

# Combine plots in a grid
wrap_plots(plots, ncol = 2)
```
We can clearly see a trend of the data leaning to the left, with the BILL_AMT histograms being almost identical, and most of the data being around 0. Let's check the boxplots next.

```{r, fig.width=14, fig.height=15}
plots <- lapply(quant_vars, function(var){
  ggplot(data, aes(x = .data[[var]])) +
    geom_boxplot() +
    theme_classic()
})
wrap_plots(plots, ncol = 2)
```
We can see a clear asymmetry, and quite a few outliers across the board.

```{r}
skewness_values = sapply(data[, quant_vars], skewness, na.rm = TRUE)
skewness_df <- data.frame(
  Variable = quant_vars,
  Skewness = skewness_values,
  Interpretation = ifelse(abs(skewness_values) < 0.5, "Symmetric",
                   ifelse(abs(skewness_values) < 1, "Moderately Skewed", "Highy Skewed"))
)
skewness_df
```
With this we can safely assume the data is negatively skewed across all quantitative variables.
```{r}
# Bar plot of skewness values
ggplot(skewness_df, aes(x = reorder(Variable, Skewness), y = Skewness, fill = Skewness > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Skewness of Quantitative Variables",
       x = "Variable", y = "Skewness") +
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "coral"), 
                    labels = c("Negative", "Positive"),
                    name = "Direction")
```
```{r}
# Detect outliers for all quantitative variables
outliers_list <- lapply(data[, quant_vars], detect_outliers_iqr)

# Count outliers per variable
outlier_counts <- sapply(outliers_list, sum, na.rm = TRUE)
outlier_percent <- outlier_counts / nrow(data) * 100

outlier_summary <- data.frame(
  Variable = quant_vars,
  Outliers = outlier_counts,
  Percent = round(outlier_percent, 2)
)
print(outlier_summary)
```
The percentages of outliers for LIMIT_BAL and AGE are healthy. As for the outlier percentages for BILL_AMT1 through BILL_AMT6, the values are concerning. Let's analyse the results
```{r}
# Commencons par les histogrammes
for (var in quant_vars) {
  p <- ggplot(data, aes(x = .data[[var]])) +
    geom_histogram(aes(y = after_stat(density))) +
    geom_density() +
    theme_dark()
  print(p)
}
```

#Vérification de l'équilibre des data
```{r}
ggplot(data, aes(default.payment.next.month)) +
  geom_bar(aes(y = after_stat(count / sum(count) * 100), fill = default.payment.next.month)) +
  scale_fill_manual(values = c("Non-Defaut" = "green", "Defaut" = "red")) +
  labs(y = "Percentage (%)", x = "Payment Defaut Status") +
  guides(fill = "none")  # Remove legend if you don't need it
```

```{r}
percent <- prop.table(table(data$default.payment.next.month)) * 100
percent
```

On remarque que 22.12% des clients sont en défaut, puisque c'est supérieur à 5%, on peut dire que les data sont équilibrés.







